sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-01 doc/paper.Rmd
suggested edits to VAST para

# Spatially varying coefficients applied to citizen-science owl counts

@@

@@

```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
f <- here::here("figs", "timing2-logx.pdf")
if (!file.exists(f)) {
  download.file("https://www.dropbox.com/s/3aco85mizuplz7z/timing2-logx.pdf?dl=1", 
    destfile = f)
}
knitr::include_graphics(f)
```

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1). 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.

>>>>

# Spatially varying coefficients from citizen-science count data

@@

<!-- \clearpage  -->

@@

```{r timing, fig.cap="(ref:timing-fig)", out.width="3.5in", fig.align='center'}
f <- here::here("figs", "timing2-logx-PE.pdf")
if (!file.exists(f)) {
  download.file("https://www.dropbox.com/s/3aco85mizuplz7z/timing2-logx.pdf?dl=1", 
    destfile = f)
}
knitr::include_graphics(f)
```

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1). 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in the user interfaces.
<!-- could add: 'and functionality' to the above if next sentence deemed redundant. -->
<!-- The major difference lies in VAST's considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF). -->
Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
From a fisheries perspective, sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications or support delta/hurdle models.
<!-- I think either this final sentence or the 3rd sentence are a bit redundant. I suggest removing the 3rd one here. -->
